`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 


# The First Appendix

<!--
If you feel it necessary to include an appendix, it goes here.
-->

This first appendix includes all of the R chunks of code that were hidden throughout the document (using the `include = FALSE` chunk tag) to help with readibility and/or setup.

**In the main Rmd file**

```{r ref.label='include_packages', results='hide', echo = TRUE}
```

**In Chapter \@ref(ref-labels):**

```{r ref.label='include_packages_2', results='hide', echo = TRUE}
```

# List of abbreviations

Poledwid: Width of poll edge (g)

Lefedwid: Max width of left side edge (mm)

Ritedwid: Max width of right side edge (mm)

Blaedwid: Width of blade edge (mm)

Bladang: Angle at blade edge (degrees)

Grifropl: Area of grinding on poll, front face (cm2)

Grifroby: Area of grinding on body, front face (cm2)

Grifrobl: Area of grinding on blade, front face  (cm2)

Gribapl: Area of grinding on poll, back face (cm2)

Gribaby: Area of grinding on body, back face (cm2)

Gribabl: Area of grinding on blade, back face (cm2)

Variables beginning with POL, REF and COR are the same as GRI measurements but instead of 
grinding replace with polish, reworking and cortex respectively, and also in cm2

Blacon: Concavity of blade (mm)

Blasin: Blade sinuosity (mm)

Edcon: Right edge concavity (mm)

Edsin: Right edge sinuosity (mm)

Maxthick: Maximum thickness (mm)

Polthick: Thickness of poll (mm)

Blathick: Thickness of blade (mm)

Lefthick: Thickness left of centre, opposite point of max thickness (mm)

Rithick: Same as LEFTHICK, but right of centre (mm)

Topthick: Thickness of top half (mm)

Bothick: Thickness of bottom half (mm)

Maxleng: Maxmimum length (mm)

Poleng: Poll length (mm)

Blaleng: Blade length (mm)

Maxwid: Maximum width (mm)

Polwid: Poll width (mm)

Bladwid: Blade width (mm)

Lengmaxt: Length from blade to point of max thickness (mm)

Lengmaxw: Length from blade to point of max width (mm)

Gripoled: Length of poll edge ground (mm)

Lonpoled: Total length of poll edge (mm)

Grinside: Extent of grinding on left side edge (mm)

Longside: Length of left side edge (mm)

Areapoll: Area of poll (cm2)

Areabody: Area of body (cm2)

Areablal: Area of left segment of blade (cm2)

Areablar: Area of right segment of blade (cm2)

Areablad: Total area of blade (cm2)