---
author: 'GTNV0'
date: 'November 2020'
division: 'Computational Archaeology'
advisor: 'Andrew Bevan'
altadvisor: 'Mark Altaweel'
department: 'Institute of Archaeology'
title: 'Exploring shape and size in British Neolithic Axes using old data: a comparison of methods'
knit: "bookdown::render_book"
site: bookdown::bookdown_site

output: 
  thesisdown::thesis_pdf: default

abstract: '`r if(knitr:::is_latex_output()) paste(readLines(here::here("prelims", "00-abstract.Rmd")), collapse = "\n  ")`'
acknowledgements: |
  Many thanks to Andrew Bevan for providing mentoring and introducing me to this dataset, to Michael Pitts for creating it, and to the Implement Petrology Group for scanning it and making it publicly available.
dedication: |
  You can have a dedication here if you wish. 
preface: |
  This is an example of a thesis setup to use the reed thesis document class
  (for LaTeX) and the R bookdown package, in general.
  
bibliography: bib/thesis.bib

csl: csl/harvard-institute-of-education-2.csl

lot: true
lof: true
header-includes:
- \usepackage[all]{nowidow}
---


```{r eval=!knitr::is_latex_output(), child=here::here("prelims", "00--prelim.Rmd")}

```

```{r eval=!knitr::is_latex_output(), child=here::here("prelims", "00-abstract.Rmd")}

```

# Introduction {.unnumbered}

This dissertation follows two main lines of investigation: how data can be extracted from legacy formats for use with computational methods, and whether size, shape or both can cluster Neolithic axes from Britain. The author firstly takes special interest in analogue datasets as they may not have been used for years or with recent techniques, and also because they pose unique challenges in data extraction and cleanup. Extracting old archaeological data could help to give new insight into the past through use with modern methods, and helps to highlight that much paper-based information remains unused ever, or in recent years, which could be tapped into. Secondly, the differing morphology of British Neolithic axes is of particular interest as these tools are so ubiquitous in the archaeological record, yet the reasons behind their designs are still relatively ambiguous. Further, quantifying traits of artefacts to aid in grouping as an alternative to classical typological methods is of particular appeal for approaching the archaeological problem domain inferentially. In light of and in combination with recent research on British Neolithic axes, investigating this seldom-pursued aspect could help to reveal more about these tools and their manufacturers from a period with no surviving writing which lasted thousands of years.

The data used in this project is a paper-based dataset compiled by Mike Pitts in the 1970s, scanned to images in 2016, which contains technical drawings of British Neolithic axes and record sheets cataloguing their measurements in addition to other information such as their find places and materials. Before any data extraction, the literature of both the archaeological and methodological contexts relevant to this dissertation are reviewed in order to place this research into perspective.  Next, to extract the analogue data, transcriptions of the record sheets were crowdsourced using the MicroPasts online platform, and the drawing images were processed using Python’s OpenCV library into a format compatible with the Momocs R package for outline morphometry. After the data was digitised, the two sets were then subjected to similar separate analyses. The measurements from the record sheet data were explored first with density and violin plots and then analysed with principal components analysis and k-means clustering, whilst the processed drawings were elliptical Fourier-transformed and then subjected to the same multivariate analyses. The results of both sets of k-means clustering were then compared by plotting the find places of the axes from each on a map of Britain to see which clusters persisted geographically, in addition to finding the similarity in axe membership between the two sets of clusters. All data extraction methods, analyses, and the building of this dissertation PDF were made reproducible using a Makefile, Docker, renv for R and venv for Python. The implications of the results from both sets of analyses and their comparisons were then discussed in light of the initial literature review.